---
title: "Peer Assessment 1 - home work"
output: html_document
---

We first introduce libraries **dplyr**, **lubridate** and **lattice**, and then load the data:

```{r}
library(dplyr)
library(lubridate)
library(lattice)
mydata <- read.csv("viikkoprojekti/repdata_data_activity/activity.csv", na.strings="NA")

```

Then we modify text based date as POSIX date field:

```{r}
mydata$date <- as.Date(mydata$date,"%Y-%m-%d")

```

## What is mean total number of steps taken per day?

We create a new data frame called **sumbyday** to group data by date, and summarise steps by date
  
```{r}
 sumbyday <- mydata %>%
 filter(!is.na(steps)) %>%
 group_by(date) %>%
 summarise(totalsteps=sum(steps)) 
```

Then we calculate mean and average of steps:

```{r}
 mean_of_steps = mean(sumbyday$totalsteps)
 median_of_steps = median(sumbyday$totalsteps)
 mean_of_steps
 median_of_steps
```

We print the histogram based on sumbydat data frame:

```{r}
hist(sumbyday$totalsteps, xlab="Total Steps", main="Histogram - Total Number of Steps")

```

##What is the average daily activity pattern?

We group data by interval and calculate sum of steps and average of steps:

```{r}
byinterval <- mydata %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarise(totalsteps=sum(steps),average.steps=mean(steps)) 
```

Time series for Total number of steps per 5-minutes interval:

```{r}
with(byinterval,plot(byinterval$interval, byinterval$average.steps, type="l",main="Total number of steps per 5-minutes interval", xlab="Interval",ylab="Sum of steps"))

```


Then we calculcate maximum number of steps in a five minute interval


```{r}

maxint <- max(byinterval$totalsteps)
byinterval[maxint, ]
```


## Imputing missing values

We Calculate and report the total number of missing values in the dataset


```{r}
nr.na <- sum(is.na(mydata$steps))
nr.na
```

Replace missing values from the mean of the 5- minute interval

```{r}
mydata.withoutNA <- mydata %>%
  group_by(interval) %>%
  mutate (steps = ifelse(is.na(steps), mean(steps,na.rm=TRUE), steps))

```

Then we calculate sum of steps in each day for that mutated data set.

```{r}
sumbyday.withoutNA <- mydata.withoutNA %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarise(totalsteps=sum(steps))

```

We calculate and report mean of steps and median of steps of mutated data set


```{r}


mean_of_steps_withouthNA = mean(sumbyday.withoutNA$totalsteps)
median_of_steps.WithouthNA = median(sumbyday.withoutNA$totalsteps)
mean_of_steps_withouthNA
median_of_steps.WithouthNA

```

and finally print histogram of sum of steps without missing values


```{r}
hist(sumbyday.withoutNA$totalsteps, xlab="Total Steps", main="Task 1: Histogram - Total Number of Steps without missing values")

```

The impact of replacing NAs with mean values is that number of steps will grow (obviously), but also histogram chart will be more balanced around central point

##Are there differences in activity patterns between weekdays and weekends?

First we use function **wday()** whichreturns weekday as a number 01-07, Sunday is 1.
Then we modify the data set to contain information wether the date is a 'weekday' or 'weekend'
```{r}

mydata.withoutNA$weekday <- wday(mydata.withoutNA$date)  
mydata.withoutNA$daytype <- ifelse(mydata.withoutNA$weekday %in% c(7,1), "weekend", "weekday" )

```

now we group data by daytype and interval


```{r}
mydata.daytype <- mydata.withoutNA %>%
  group_by(daytype,interval) %>%
  summarise(total.steps = sum(steps, na.rm=TRUE), average.steps=mean(steps,na.rm=TRUE))

```

then we use lattice library xyplot function to plot both weekend and weekdays charts

```{r}


xyplot(average.steps~interval|daytype, data=mydata.daytype,type="l", main="Average Daily Activity on Weekdays and weeekends", ylab="Number of Steps per", xlab="Time Interval")


```

